---
title: "Practical Machine Learning Assignment"
author: "Andrew"
date: "22/11/2015"
output: html_document
---

# Executive Summary

The goal of your project is to predict the manner in which participants did exercise. The data for this assignment comes from the [Human Activity Recognition](http://groupware.les.inf.puc-rio.br/har) dataset. 

# Analysis

First we define helper functions to load additional packages (usePackage) and download/load CSV data (loadCSVData): 
```{r}
rm(list=ls())

usePackage<-function(p){
  # load a package if installed, else load after installation.
  # Args:
  #   p: package name in quotes
  
  if (!is.element(p, installed.packages()[,1])){
    print(paste('Package:',p,'Not found, Installing Now...'))
    install.packages(p, dep = TRUE)}
  print(paste('Loading Package :',p))
  require(p, character.only = TRUE)  
}

loadCSVData<-function(url){
  # saves a data file to a temporary location and loads into memory.
  # dir.create('tmp', showWarnings = FALSE)
  file_name <- basename(url)
  
  if (file.exists(file_name)) {
   print('Using local file.')
} else {
  print('Downloading file.')
  download.file(url, file_name, method = "curl")
}
  file_data <- read.csv(file_name, na.strings = c("", "NA", "#DIV/0!"))
  return(file_data)
}

```

## Examining Dataset

Using the helper function to download the test and training data from the url's provided: 
```{r}
train_url <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
test_url <- 'https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

train_data <- loadCSVData(train_url)
test_data <- loadCSVData(test_url)
```

and examine the data set (verbose output surpressed) :

```{r,echo=FALSE, results='hide',message=FALSE}
head(train_data)
summary(train_data$classe)
summary(test_data)
names(train_data)
```

We see the class varible in the training data contains 5 categories [A, B, C, D, E]. The objective of this analysis is to classify the test data into one of the five categories. We then clean up the data set, removing the columns that contain mostly N/A's: 

```{r}



# convert class to factor:
train_data$classe <- as.factor(train_data$classe)
test_data$classe <- NA
levels(test_data$classe) <- levels(train_data$classe)

# removing col with more than 90% N/A's: 
na_count_limit <- 0.90

na_per_col <- apply(train_data, 2, function(x) {sum(is.na(x))})
train_data <- train_data[ , which(na_per_col <  nrow(train_data)*na_count_limit)]

keep_col <- names(train_data)
test_data <- test_data[, keep_col]

# removing non relevant columns:
train_data$X <- NULL
train_data$user_name <- NULL
train_data$cvtd_timestamp <- NULL
train_data$new_window <- NULL
train_data$num_window <- NULL
train_data$raw_timestamp_part_1 <- NULL
train_data$raw_timestamp_part_2 <- NULL

test_data$X <- NULL 
test_data$new_window <- NULL
test_data$num_window <- NULL
test_data$raw_timestamp_part_1 <- NULL
test_data$raw_timestamp_part_2 <- NULL



```

## Fitting Model

We divide the training data into two sets (training and validation). 

```{r}
usePackage('caret')
set.seed(1)
train_toggle <- createDataPartition(train_data$classe, p = 0.8, list = FALSE)
training_set <- train_data[train_toggle, ]
validation_set <- train_data[-train_toggle, ]
```

And fit a random forrest model: 

```{r}
usePackage('randomForest')
set.seed(1)
model_rf <- randomForest(classe ~ .,  data=training_set, ntrees = 100, na.action = na.omit)
plot(model_rf, log="y")
```

## Examining Model Fit

Examining the model fit to the training data (by fitting the training data):

```{r}

predict_train <- predict(model_rf, training_set, type = "class")
confusionMatrix(predict_train, training_set$classe)
```

And examining the model fit using the validation data:

```{r}
predict_validation <- predict(model_rf, validation_set, type = "class")
confusionMatrix(predict_validation, validation_set$classe)
```

Reading from the above output, the cross validation accuracy is 99.62% (corrresponding to an error of 0.38%).

## Fitting Test Set

Using the random forest model to classify the test data set: 

```{r}
test_class <- predict(model_rf, test_data)
test_class
```

# Conclusions

A random forest model was used to classify the test data by class. The model cross validation accuracy was 99.52% (an error of 0.38%) and the model predicts the class varible in the test data as follows: 

 1,  2,  3,  4,  5,  6,  7,  8,  9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20,
 
 B,  A,  B,  A,  A,  E,  D,  B,  A,  A,  B,  C,  B,  A,  E,  E,  A,  B,  B,  B, 

# Submission

Preparing files for submission: 

```{r}
answers <- as.vector(test_class)
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(answers)

```

